EDIpack 2.0: A generic and interoperable Lanczos-based solver for quantum impurity problems

### 2025-05-20 01_intro.tex
Iterated introduction. Ready to update W2Dyn section with T>0 results.
Co-authored-by: Adriano Amaricci <adriano.amaricci@gmail.com>
Co-authored-by: Samuele Giuli <samuele.giuli@gmail.com>

\section{Introduction and Motivation}\label{SecIntro}

A few words about the motivations who led us to develop this
software, possible applications and advantages:

* flexibility: can address generic cases including multi-orbital,
superconducting or spin-non-conserving regimes

* zero and low finite temperatures

* direct access to (well approximated) analytic dynamical functions

* direct access to impurity Fock space quantities

bla bla bla

>>>>

\documentclass[edipack2.tex]{subfiles}
\begin{document}

\section{Introduction and Motivation}\label{SecIntro}
\fixme{A tentative Introduction scheme}

Quantum impurity models play a central role in the study of strongly correlated electron systems, capturing the coupling of a small number of localized degrees of freedom with extended non-interacting environment.   
Originally introduced to describe magnetic impurities in metallic hosts, i.e. Kondo problem, these systems progressively became powerful tools for exploring a wide range of quantum many-body phenomena. More recently, their importance extended beyond the realm of condensed matter, influencing understanding of other areas such as quantum information or ultra-cod atomic systems.

At the heart of quantum impurity models is the interaction between a localized impurity, possibly including lattice, orbital and spin states, and a surrounding bath of itinerant particles, usually represented as a conduction band of electrons. 
This setup capture many essential aspects of strong correlation including screening, decoherence and entanglement in a computationally tractable way. 
Most notably, quantum impurity problems lie at the core of dynamical mean-field theory (DMFT) where the complex behavior of the lattice systems is reduced to an effective self-consistent quantum impurity problem.  
By retaining essential local correlations, the DMFT enabled a successful effective description of the most important physical aspects of correlated materials, including Mott insulators, heavy fermion compounds, and high-temperature superconductors.

However, solving the quantum impurity problems emerging in the DMFT context accurately and efficiently remains a
critical challenge, particularly for multi-orbital systems, systems
at low temperatures or endowed with lower symmetry. During many years of development, different advanced numerical techniques have been developed together with cutting edge approaches aimed to reduce computational load.

%EDIPACK 2.0
In this work we introduce  EDIpack 2.0: a flexible, high-performance numerical library designed to address these challenges, providing an exact diagonalization (ED) solver for generic Quantum Impurity Problems. 
The library builds on the foundations of
its predecessor, featuring massively parallel Lanczos-based algorithms while including several significant extensions and 
improvements. The library now supports both zero and finite temperature calculations for different symmetries of the problem within a unified framework, making it suitable for a wide range of physical phenomena, from spin-orbit coupling effect in quantum materials to multi-orbital superconductivity. It also provides direct
access to well-approximated dynamical correlation functions across the
entire complex plane, thus enabling calculations of spectral functions, local susceptibilities, and
other response properties.

Additionally, \NAME offers access to the Fock space of
the impurity system and thus, for instance, the possibility to
evaluate the impurity reduced density matrix
(iRDM). This capability enables quantum information-inspired analyses, providing insights into the entanglement properties, subsystem purity, and quantum correlations of interacting electron systems. These
features are increasingly recognized as essential for understanding the emergent behavior in correlated matter, including quantum criticality and superconductivity.

This new version also includes support for systems with non-equivalent impurities , a critical feature to study hetero-structures, super-lattices, and disordered systems, where the local environment varies significantly across different sites. This capability makes it 
possible to explore a wide range of inhomogeneous systems, including
oxide interfaces or disordered systems, where
translational symmetry is broken.

%\paragraph{Interoperability}\\
With the growing importance of  quantum impurity models in the study of correlated materials, ensuring compatibility among different methods and developments has become a key challenge.
In this context, interoperability became a core design principle of this new updated version of \NAME. The library is implemented using modern Fortran constructs, with comprehensive C-bindings that enable integration with other programming languages. 
In particular, \NAME includes native C and C++
interface layer which enable development of third-party
extensions. A significant contribution is realized via the Python API, EDIpy, which provides a high-level interface for defining and solving impurity problems. The
Python bindings also enable integration with established scientific
libraries, such as NumPy and SciPy, facilitating data analysis and
rapid prototyping.

A key development feature in \NAME, unlocked by the introduction of Python and C/C++ API, is the direct  interfaces to the TRIQS
(Toolbox for Research on Interacting Quantum Systems)  as
well as to the W2Dynamics software, which are two of the most important and used computational tools in the field. 
This integration provides \NAME users with access to the
extensive data handling and analysis tools in the TRIQS ecosystem,
while also allowing TRIQS and W2Dynamics users to benefit from the
flexibility and performance of \NAME solver. This dual functionality
extends the capabilities of these platforms, supporting
reproducibility and cross-validation of numerical results, and sets a
new standard for interoperability in quantum impurity solvers.

%\paragraph{closure}
In this paper, we present the mathematical framework, implementation
details, and key features of \NAME, along with several illustrative
applications. We aim to demonstrate how this flexible,
high-performance solver can be used to tackle a wide range of
challenging impurity problems. 
The rest of this work has the following structure. In \secu{SecInstall} we 
briefly illustrate the overall structure of \NAME together with its dependencies, configuration and installation procedures. In the following  \secu{SecEDIpack} we present the quantum impurity problem and review the software implementation in full details. In the following \secu{SecInterop} we thoroughly discuss the C-binding interface which is at the heart of the interoperability capabilities of \NAME. In this section we also discuss 4 interface layers: i) the Python API, ii) the TRIQS interface, iii) the W2Dynamics interface and iv) a basic Julia API. 
Finally, in \secu{SecExamples} we discuss the functionalities and the capabilities of \NAME and all the discussed extensions with a series of illustrative examples.

\end{document}

### 2025-05-20 04_cbinding.tex
Iterated introduction. Ready to update W2Dyn section with T>0 results.
Co-authored-by: Adriano Amaricci <adriano.amaricci@gmail.com>
Co-authored-by: Samuele Giuli <samuele.giuli@gmail.com>

\subsection{W2Dynamics interface}\label{sSecInteropW2DYN}
\NAME is supported as alternative impurity solver by the w2dynamics DMFT package\cite{Wallerberger2019CPC}. The only prerequisite to use this option is the presence of a functioning EDIpy2 installation, which w2dynamics will attempt to import at runtime only if the user sets the appropriate configuration parameter.

>>>>

\paragraph{Anaconda}
Another option for installing the package is by using the Anaconda package 
manager on Unix/Linux systems.
The following commands will create a new {\tt conda} environment named 
`edipack' and install the most recently released version of edipack2triqs along 
with its dependencies (\NAME, EDIpy2 and the TRIQS libraries).

\begin{lstlisting}[style=mybash]
conda create -n edipack
conda activate edipack
conda install -c conda-forge -c edipack edipack2triqs
\end{lstlisting}

\subsubsection{Implementation}\label{sSecInteropTRIQSImplementation}
The programming interface of edipack2triqs is built around the singleton 
Python class {\tt EDIpackSolver} (defined in the module
{\tt edipack2triqs.solver}), whose instance represents the internal state of 
the \NAME library and exposes its functionality through a number of 
attributes and methods.

\paragraph{{\bf Constructor}}
Constructor of {\tt EDIpackSolver} accepts the Hamiltonian to be diagonalized
in the form of a second-quantized fermionic operator, the TRIQS {\tt Operator}
object described in Sec.~8.7 of Ref.~\onlinecite{Parcollet2015CPC} (impurity 
problems involving bosons are not supported in the current version). In 
addition to the Hamiltonian, four {\it fundamental operator sets}
{\tt fops\_imp\_up}, {\tt fops\_imp\_dn}, {\tt fops\_bath\_up} and
{\tt fops\_bath\_dn} must be provided.
Each of the sets contains pairs of labels {\tt (b,j)} carried
by the operators $c^\dagger_{b,i} / c_{b,i}$ corresponding to either impurity
or bath electronic degrees of freedom with a certain spin projection.
Having these two crucial pieces of information, the constructor initializes
the underlying Fortran library and automatically selects the appropriate
{\tt ed\_mode=normal,superc,nonsu2} (Sec.~\ref{sSecQNs}) and
{\tt bath\_type=normal,hybrid,general} (Sec.~\ref{sSecBath}). Besides, the
constructor accepts a vast array of keyword arguments that allow for fine-tuning
the diagonalization process. Among others, these include the ED algorithm
selection, the quantum numbers to be used, parameters of the Krylov space,
the spectrum cutoff and various tolerance levels.

\paragraph{{\bf Method {\tt solve}}}
The method {\tt solve()} of {\tt EDIpackSolver} calls
{\tt edipy2.global\_env.solve()} and therefore performs the bulk of
calculations. It accepts the inverse temperature $\beta$ required to calculate
the expectation value of physical observables, and parameters of energy grids
for Green's function calculations.

\paragraph{{\bf Input parameters}}
It is possible to read off and change parameters of the Hamiltonian between
successive calls to {\tt solve()} via respective attributes of {\tt 
EDIpackSolver}. Note, however, that the changes that would necessitate a change of {\tt ed\_mode} or {\tt bath\_type} are disallowed. The relevant attributes
are the following.
\begin{itemize}
    \item {\tt nspin} --- the number of non-equivalent spin projections 
          (read-only).
    \item {\tt norb} -- the number of impurity orbitals (read-only).
    \item {\tt hloc} --- matrix $h^0_{\alpha\beta\s\s'}$ of
          the quadratic impurity Hamiltonian (\ref{Himp}).
    \item {\tt Uloc}, {\tt Ust}, {\tt Jh}, {\tt Jx}, {\tt Jp} ---
          parameters $U, U', J, J_X \text{ and } J_P$ of the impurity
          interaction term $\hat{H}^\mathrm{int}_\mathrm{imp}$ (\ref{Hint}).
    \item {\tt bath} --- an object representing the bath. Depending on the
          bath type selected upon solver object construction, this object is an
          instance of {\tt BathNormal}, {\tt BathHybrid} or {\tt BathGeneral}
          (all three classes are members of the module
          {\tt EDIpackSolver.bath}). The bath objects support basic arithmetic
          operations so that a mixing scheme within a DMFT calculation can be
          easily implemented. The way to access individual bath parameters is
          specific to each of the three classes, and is described in detail in the       \href{https://krivenko.github.io/edipack2triqs/documentation.html\#module-edipack2triqs.bath}{online API reference}.
\end{itemize}

\paragraph{{\bf Calculation results}}
After a successful invocation of {\tt solve()}, one can extract results of the 
calculation from the following attributes.
\begin{itemize}
    \item {\tt e\_pot}, {\tt e\_kin} --- thermal average of the potential 
          (interaction) and kinetic energy respectively.
    \item {\tt densities}, {\tt double\_occ} --- lists of average densities and
          double occupancies of impurity orbitals.
    \item {\tt magnetization} --- Cartesian components of the average impurity  
          magnetization vectors, one row per orbital.
    \item {\tt superconductive\_phi} --- Matrix of impurity superconductive
          order parameters
          $\phi_{\alpha\beta} = \langle c_{\alpha\up} c_{\beta\down}\rangle$
          in orbital space. {\color{red} Is it defined elsewhere in the text?}
    \item {\tt g\_iw}, {\tt g\_an\_iw} --- Normal and anomalous components of
          the Matsubara impurity Green's function.
    \item {\tt Sigma\_iw}, {\tt Sigma\_an\_iw} --- Normal and anomalous
          components of the Matsubara impurity self-energy.
    \item {\tt g\_w}, {\tt g\_an\_w} --- Normal and anomalous components of
          the real-frequency impurity Green's function.
    \item {\tt Sigma\_w}, {\tt Sigma\_an\_w} --- Normal and anomalous
          components of the real-frequency impurity self-energy.
\end{itemize}

The Green's functions are returned as TRIQS {\tt BlockGf} containers
with names of the individual blocks determined from the block labels {\tt 'b'}
found in {\tt fops\_imp\_up}, {\tt fops\_imp\_dn} and from solver's
{\tt ed\_mode}. The Matsubara and real-frequency meshes of the TRIQS GF 
containers (see Sec.~8.2 of Ref.~\onlinecite{Parcollet2015CPC}) are constructed
according to the parameters passed to {\tt solve()}.
The anomalous components of the Green's functions and self-energies are only 
available when anomalous bath terms are present in the Hamiltonian.
If this is not the case, an attempt to access these attributes results in a
{\tt RuntimeError}.

\paragraph{{\bf Bath parameter fitting}}
The method {\tt EDIpackSolver.chi2\_fit\_bath()} is essentially a wrapper around
{\tt edipy2.global\_env.chi2\_fitgf()}. It accepts the function to fit (either
the hybridization function or the Weiss field) in the {\tt BlockGf} format,
and returns the parameter fit result as a {\tt Bath*} object along with a
{\tt BlockGf} representation of the fitted function. In the superconducting
case ({\tt ed\_mode=superc}), this method accepts and returns pairs of
the {\tt BlockGf} containers corresponding to the normal and anomalous 
components of the quantities in question.

\subsection{W2Dynamics interface}\label{sSecInteropW2DYN}
\NAME is supported as alternative impurity solver by the w2dynamics DMFT package\cite{Wallerberger2019CPC} available on \href{https://github.com/w2dynamics/w2dynamics}{GitHub}. The only prerequisite to use this option is the presence of a functioning EDIpy2 installation, which w2dynamics will attempt to import at runtime only if the user sets the appropriate configuration parameter.
